---
title: "Profit Projection for Timeless Transport Models"
output: html_document
date: "2025-03-18"
---

# Summary

Timeless Transport Models wants to know the expected profit going into the second half of 2005. Using production and sales data to:

1)  Estimate the unit production costs using linear regression.

2)  Find the average profit per unit for each product line.

3)  Forecast of monthly sales from June 2005 to December 2005 using time series modeling.

4)  Compute expected profit calculation to find the total profit.

Findings:

TImeless Transport Models is expected to make \$1,113,052.04 in profit from June 2005 to December 2005.

# Classic Cars Sales Forecast

```{r}
library(tidyverse)
sales <- read.csv("data/sales_data_sample.csv")

# Add an order_month column to sales that is the month the order took place. 
sales$order_month = floor_date(as.Date(sales$ORDERDATE, "%m/%d/%Y"), unit="month")

# Filter the sales data for Classic Cars product line
cc_sales <- sales[sales$PRODUCTLINE == "Classic Cars",]

# Summarize the total quantity ordered for each month
cc_monthly_sales <- cc_sales %>%
  group_by(order_month) %>%
  summarise(qty = sum(QUANTITYORDERED)) 

# Make the time series object
cc_ts <- ts(cc_monthly_sales$qty, start=c(2003, 1), frequency=12)
```

```{r}
library(forecast)
#Seasonal Naive method to forecast monthly sales
# cc_forecast <- forecast(snaive(cc_ts), h = 7)
# print(cc_forecast)
# autoplot(cc_forecast)

snaive(cc_ts, 7)
autoplot(snaive(cc_ts, 7))
```

# Ships Sales Forecast

```{r}
# Filter the sales data for the Ships product line
ships_sales <- sales[sales$PRODUCTLINE == "Ships",]

# Summarize the total quantity ordered for each month
ships_monthly_sales <- ships_sales %>%
  group_by(order_month) %>%
  summarise(qty = sum(QUANTITYORDERED)) 

# Make the time series object
ships_ts <- ts(ships_monthly_sales$qty, start=c(2003, 1), frequency=12)
```

```{r}
#Seasonal Naive method to forecast monthly sales
snaive(ships_ts, 13)
autoplot(snaive(ships_ts, 13))

```

# Motorcycles Sales Forecast

```{r}
# Filter the sales data for the Motorcycles product line
motorcycles_sales <- sales[sales$PRODUCTLINE == "Motorcycles",]

# Summarize the total quantity ordered for each month
motorcycles_monthly_sales <- motorcycles_sales %>%
  group_by(order_month) %>%
  summarise(qty = sum(QUANTITYORDERED)) 

# Make the time series object
motorcycles_ts <- ts(motorcycles_monthly_sales$qty, start=c(2003, 1), frequency=12)
```

```{r}
#Seasonal Naive method to forecast monthly sales
snaive(motorcycles_ts, 9)
autoplot(snaive(motorcycles_ts, 9))
```

# Trucks and Buses Sales Forecast

```{r}
# Filter the sales data for the Trucks and Buses product line
tb_sales <- sales[sales$PRODUCTLINE == "Trucks and Buses",]

# Summarize the total quantity ordered for each month
tb_monthly_sales <- tb_sales %>%
  group_by(order_month) %>%
  summarise(qty = sum(QUANTITYORDERED)) 

# Make the time series object
tb_ts <- ts(tb_monthly_sales$qty, start=c(2003, 1), frequency=12)
```

```{r}
#Seasonal Naive method to forecast monthly sales
snaive(tb_ts, 10)
autoplot(snaive(tb_ts, 10))
```

# Planes Sales Forecast

```{r}
# Filter the sales data for the Planes product line
planes_sales <- sales[sales$PRODUCTLINE == "Planes",]

# Summarize the total quantity ordered for each month
planes_monthly_sales <- planes_sales %>%
  group_by(order_month) %>%
  summarise(qty = sum(QUANTITYORDERED)) 

# Make the time series object
planes_ts <- ts(planes_monthly_sales$qty, start=c(2003, 1), frequency=12)
```

```{r}
#Seasonal Naive method to forecast monthly sales
snaive(planes_ts, 13)
autoplot(snaive(planes_ts, 13))
```

# Vintage Cars Sales Forecast

```{r}
# Filter the sales data for the Vintage Cars product line
vc_sales <- sales[sales$PRODUCTLINE == "Vintage Cars",]

# Summarize the total quantity ordered for each month
vc_monthly_sales <- vc_sales %>%
  group_by(order_month) %>%
  summarise(qty = sum(QUANTITYORDERED)) 

# Make the time series object
vc_ts <- ts(vc_monthly_sales$qty, start=c(2003, 1), frequency=12)
```

```{r}
#Seasonal Naive method to forecast monthly sales
snaive(vc_ts, 7)
autoplot(snaive(vc_ts, 7))
```

# Average Profit For Each Product Line

Classic.Cars: 113.002 - 78.1465 = \$34.8555

Ships: 83.02 - 15.3710 = \$67.649

Motorcycles: 107.536 - 74.2201 = \$33.3159

Trucks.and.Buses: 115.265 - 90.0914 = \$25.1736

Planes: 94.887 - 76.8948 = \$17.9922

Vintage.Cars: 94.036 - 60.3941 = \$33.6419

# Total Forecast Units For Each Product Line

Classic.Cars: 742 + 1263 + 1885 + 799 + 1786 + 3669 + 1006 = 11,150 units

Ships: 303 + 320 + 102 + 545 + 76 + 303 + 294 = 1,943 units

Motorcycles: 675 + 430 + 427 + 1403 + 268 + 377 + 429 = 4,009 units

Trucks.and.Buses: 351 + 415 + 750 + 1052 + 569 + 337 + 356 = 3,830 units

Planes: 681 + 175 + 270 + 883 + 496 + 383 + 414 = 3,303 units

Vintage.Cars: 613 + 481 + 849 + 793 + 1385 + 2305 + 818 = 7,244 units

# Expected Profit = Average Profit \* Total Forecast Units

Classic.Cars: 31.583920 \* 11,150 = \$352,117.83

Ships: 71.999960 \* 1,943 = \$139,895.92

Motorcycles: 24.917439 \* 4,009 = \$99,894.01

Trucks.and.Buses: 13.608588 \* 3,830 = \$52,120.924

Planes: 7.788175 \* 3,303 = \$25,724.34

Vintage.Cars: 23.656627 \* 7,244 = \$171,368.606

# Total Profit

Total Profit: $841,164.48

# Analysis

Although for the initial time series data we each used different methods to model projected sales, there was less clarity behind the multivariate nominal analysis model, and several projections for individual product lines had dramatically wider ranges. While it is likely that these estimates are more accurate and capture the statistical reality of the data more cleanly, due to the fairly consistent observed seasonal sales behavior in the data, for ease of recommendation and presentation, we chose to go with a seasonal naive model for the profit projections, as when explaining where the results came from (a simpler combination of seasonal trends and estimated growth trend), the impact can be focused on the data and what actions can be taken, rather than uncertainty regarding specific product lines.

# Conclusion

Our company's projected profit for the rest of the year can be broken down into the individual product lines to better understand product and factory performance. Over the next seven months we are projected to generate $841,164.48 USD in profit. Our application of multivariate nominal analysis may allow us a deeper view into product line uncertainty, as a more complex analysis reveals some product lines have far narrower projection ranges, and therefore are more likely to follow the seasonal and full trend. While solely focusing on these product lines is not recommended, it may serve to prioritize them in case of financial production limitations, and may indicate a need for more consistent or robust advertising around the more uncertain product lines in the pre holiday season. 

Classic.Cars: $352,117.83

Ships: $139,895.92

Motorcycles: $99,894.01

Trucks.and.Buses: $52,120.924

Planes: $25,724.34

Vintage.Cars: $171,368.60

# Total Profit

Total Profit: $841,164.48

```{r}
profit_data <- data.frame(
  Category = c("Classic Cars", "Ships", "Motorcycles", "Trucks and Buses", "Planes", "Vintage Cars"),
  Profit = c(352117.83, 139895.92, 99894.01, 52120.924, 25724.34, 171368.60)
)

total_profit <- sum(profit_data$Profit)

ggplot(profit_data, aes(x = "", y = Profit, fill = Category)) +
  geom_bar(stat = "identity", width = 0.5) +
  labs(title = "Total Profit by Category",
       y = "Profit ($)",
       x = "") +f
  geom_text(aes(x = "", y = total_profit + 50000, label = paste("Total Profit: $", format(total_profit, big.mark = ","))),
            size = 6, fontface = "bold") +
  theme(legend.position = "right")

unit_data <- data.frame(
  Category = c("Classic Cars", "Ships", "Motorcycles", "Trucks and Buses", "Planes", "Vintage Cars"),
  Units_Sold = c(11150, 1943, 4009, 3830, 3303, 7244)
)

# Create a bar plot
ggplot(unit_data, aes(x = Category, y = Units_Sold, fill = Category)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Total Units Sold by Category",
       x = "Category",
       y = "Units Sold") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")
```





# Multivariate nominal analysis
The following data shows some avenues for further investigations, notably the greater variation in projections between product lines when more factors are accounted for and a more robust regression model is applied

```{r}
sales <- read.csv("data/sales_data_sample.csv")

sales$order_month = floor_date(as.Date(sales$ORDERDATE, "%m/%d/%Y"), unit="month")

```

Breaking up into product lines, Ships, Vintage Cars, Classic Cars, Planes, Trains, Motorcycles, Trucks and Busses
And projecting sales by month, using additive seasonality
SHIPS

```{r}
sales_ships <- sales[sales$PRODUCTLINE == 'Ships',]
monthly_sales_ships <- sales_ships %>%
  group_by(order_month) %>%
  summarise(qty = sum(QUANTITYORDERED)) 

monthly_sales_ships_ts <- ts(monthly_sales_ships$qty, start=c(2003, 1), frequency=12)
monthly_sales_ships_ts

m_n_a_ships <- ets(monthly_sales_ships_ts, model = "MNA")
m_n_a_ships

forecast_ships <- forecast(m_n_a_ships, 7)

autoplot(forecast(m_n_a_ships, 7)) 

accuracy(m_n_a_ships)
forecast_ships
```

VINTAGE CARS
```{r}
sales_vintage_cars <- sales[sales$PRODUCTLINE == 'Vintage Cars',]
monthly_sales_vintage_cars <- sales_vintage_cars %>%
  group_by(order_month) %>%
  summarise(qty = sum(QUANTITYORDERED)) 

monthly_sales_vintage_cars_ts <- ts(monthly_sales_vintage_cars$qty, start=c(2003, 1), frequency=12)
monthly_sales_vintage_cars_ts

m_n_a_vintage_cars <- ets(monthly_sales_vintage_cars_ts, model = "MNA")
m_n_a_vintage_cars

forecast_vintage_cars <- forecast(m_n_a_vintage_cars, 7)

autoplot(forecast(m_n_a_vintage_cars, 7)) 

accuracy(m_n_a_vintage_cars)
forecast_vintage_cars
```


CLASSIC CARS
```{r}
sales_classic_cars <- sales[sales$PRODUCTLINE == 'Classic Cars',]
monthly_sales_classic_cars <- sales_classic_cars %>%
  group_by(order_month) %>%
  summarise(qty = sum(QUANTITYORDERED)) 

monthly_sales_classic_cars_ts <- ts(monthly_sales_classic_cars$qty, start=c(2003, 1), frequency=12)
monthly_sales_classic_cars_ts

m_n_a_classic_cars <- ets(monthly_sales_classic_cars_ts, model = "MNA")
m_n_a_classic_cars

forecast_classic_cars <- forecast(m_n_a_classic_cars, 7)

autoplot(forecast(m_n_a_classic_cars, 7)) 

accuracy(m_n_a_classic_cars)
forecast_classic_cars
```


PLANES
```{r}
sales_planes <- sales[sales$PRODUCTLINE == 'Planes',]
monthly_sales_planes <- sales_planes %>%
  group_by(order_month) %>%
  summarise(qty = sum(QUANTITYORDERED)) 

monthly_sales_planes_ts <- ts(monthly_sales_planes$qty, start=c(2003, 1), frequency=12)
monthly_sales_planes_ts

m_n_a_planes <- ets(monthly_sales_planes_ts, model = "MNA")
m_n_a_planes

forecast_planes <- forecast(m_n_a_planes, 7)

autoplot(forecast(m_n_a_planes, 7)) 

accuracy(m_n_a_planes)
forecast_planes
```


TRAINS
```{r}
sales_trains <- sales[sales$PRODUCTLINE == 'Trains',]
monthly_sales_trains <- sales_trains %>%
  group_by(order_month) %>%
  summarise(qty = sum(QUANTITYORDERED)) 

monthly_sales_trains_ts <- ts(monthly_sales_trains$qty, start=c(2003, 1), frequency=12)
monthly_sales_trains_ts

m_n_a_trains <- ets(monthly_sales_trains_ts, model = "MNA")
m_n_a_trains

forecast_trains <- forecast(m_n_a_trains, 7)

autoplot(forecast(m_n_a_trains, 7)) 

accuracy(m_n_a_trains)
forecast_trains
```


MOTORCYCLES
```{r}
sales_motorcycles <- sales[sales$PRODUCTLINE == 'Motorcycles',]
monthly_sales_motorcycles <- sales_motorcycles %>%
  group_by(order_month) %>%
  summarise(qty = sum(QUANTITYORDERED)) 

monthly_sales_motorcycles_ts <- ts(monthly_sales_motorcycles$qty, start=c(2003, 1), frequency=12)
monthly_sales_motorcycles_ts

m_n_a_motorcycles <- ets(monthly_sales_motorcycles_ts, model = "MNA")
m_n_a_motorcycles

forecast_motorcycles <- forecast(m_n_a_motorcycles, 7)

autoplot(forecast(m_n_a_motorcycles, 7)) 

accuracy(m_n_a_motorcycles)
forecast_motorcycles
```


TRUCKS AND BUSES
```{r}
sales_trucks_and_buses <- sales[sales$PRODUCTLINE == 'Trucks and Buses',]
monthly_sales_trucks_and_buses <- sales_trucks_and_buses %>%
  group_by(order_month) %>%
  summarise(qty = sum(QUANTITYORDERED)) 

monthly_sales_trucks_and_buses_ts <- ts(monthly_sales_trucks_and_buses$qty, start=c(2003, 1), frequency=12)
monthly_sales_trucks_and_buses_ts

m_n_a_trucks_and_buses <- ets(monthly_sales_trucks_and_buses_ts, model = "MNA")
m_n_a_trucks_and_buses

forecast_trucks_and_buses <- forecast(m_n_a_trucks_and_buses, 7)

autoplot(forecast(m_n_a_trucks_and_buses, 7)) 

accuracy(m_n_a_trucks_and_buses)
forecast_trucks_and_buses
```

Summary of forecasts for sales of Ships, Vintage Cars, Classic Cars, Planes, Trains, Motorcycles, Trucks and Busses for June 2005 - December 2005
```{r}
summary_forecast <- print(forecast_ships)
summary_forecast$`Point Forecast` <- summary_forecast$`Lo 80` <- summary_forecast$`Hi 80` <- summary_forecast$`Lo 95` <- summary_forecast$`Hi 95` <- NULL
summary_forecast$ships <- print(forecast_ships)[, 1]
summary_forecast$trucks_and_buses <- print(forecast_trucks_and_buses)[, 1]
summary_forecast$vintage_cars <- print(forecast_vintage_cars)[, 1]
summary_forecast$classic_cars <- print(forecast_classic_cars)[, 1]
summary_forecast$planes <- print(forecast_planes)[, 1]
summary_forecast$trains <- print(forecast_trains)[, 1]
summary_forecast$motorcycles <- print(forecast_motorcycles)[, 1]
summary_forecast
```
